### 2016-11-27 dissertation/recha.tex
dissertation v2:part recha

\chapter{Word-level Traversal of Finite State Machines using Algebraic Geometry}
\label{ch:reacha}
Reachability analysis is a basic component of sequential circuit verification, esp. for 
formal equivalence checking and model checking techniques. Concretely, in modern synthesis 
tools, in order to improve various performance indicators such as latency, clock skew or power
density, some major refactoring and attachments are made on original designs. 
Those modifications may introduce malfunctions or glitches to the whole logic. 
In localized simulation or formal verification (esp. equivalence checking), the modifications may be 
denied since the malfunctions or glitches are considered as ``faults" in this circuit.
However, if the circuit behavior is carefully investigated, it may come to a verdict that 
those ``faults" will never be activated/excitated during a restricted execution starting 
from legal initial states and with legal inputs. Thus we will call those ``faults" as 
{\bf spurious faults}, since they will not affect the circuit's normal behavior.

Almost all practical sequential logic components can be modeled as finite state machines (FSMs). 
If we apply constrains upon the machine to make it start from designated initial states and 
take in specific legal inputs, a set of reachable states can be derived. 
As long as the ``faults" can be modeled as ``bad states", we can judge whether they are 
spurious faults by checking if they sit in the reachable states. From the spurious fault validation 
perspective, reachability analysis is a must when developing full set of sequential circuit verification
techniques.

There are quite a few methods to perform reachability checking on FSMs. One among them is 
state space traversal. Conventionally the algorithm is based on bit-level techniques such as
binary decision diagrams (BDDs) and Boolean logic. We propose a new traversal algorithm on word-level,
which bring critical advantages. In this chapter the approach will be described and discussed in depth, 
with examples and experiments showing its feasibility when applied on general circuit benchmarks.

\section{Motivation}
The motivation of this part of my research consists of many inspirations and observations.
In this dissertation, we will summarize those major ones into 3 topics. First, we start from 
a sequential circuit partial scan paper to discuss the inspirations of our work as well as some 
difficulties to overcome. Secondly, we state the observation that word-level description is 
increasingly important and common in characterizing the data flow on modern large-size datapaths. 
Last but not least, we give instances that some prerequisites for supporting word-level techniques 
are already available. Thus we cover both the necessity and sufficiency of developing such 
a word-level traversal technique in our work.

\subsection{Limitations of Full Scan Algorithms}
The inspiration of this research mainly comes from a journal paper \cite{KallaPartialScan}. 
In that paper, the author proposed an traversal algorithm using concept ``implicit state enumeration".
Concretely, the algorithm is written as follows:

\begin{algorithm}[H]
\SetAlgoNoLine
\LinesNumbered
 \KwIn{Sequential circuit, number of registers to scan}
 \KwOut{Scan registers listed in decreasing order of their non-controllability}

  $from^0 = reached = S^0$\;
  $i = 0$\;
  \While{TRUE}
  {
  	$i++$\;
	$to^i = $IMAGE$(\Delta,from^{i-1})$\;
	$new^i = to^i \cdot \overline{reached}$\;
	\For{each state variable $r_j$}
	{
		record\_if\_transitions\_present\_or\_missing$(r_j,new^i)$\;
		compute\_degree\_of\_unsettability$(r_j,new^i)$\;
	}
	\If{$new^i == 0$}
	{
		break\;
	}
	$reached = reached + new^i$\;
	$from^i = new^i$\;
  }\
  \tcc{\ \ \ \ \ \ \ \ \ \ \ \ \ \ FSM traversal completed}\
  \For{each state variable $r_j$}
  {
	\If{missing transition for $r_j$}
	{
		scan state variable $r_j$\;
	}
  }
  $illegal\_states = $bdd\_complement$(reached)$\;
  \For{each state variable $r_j$}
  {
	compute\_degree\_of\_unateness$(r_j,illegal\_states)$\;
	non-controllability$(r_j)=$degree\_of\_unsettability$(r_j)+$degree\_of\_unateness$(r_j)$\;
  }
  order state variables in terms of their non-controllabilities;\tcc{\ Sorting}\
  output the required scan registers\;
%\Return{$from_k(R_{final})$}
\caption {SIMPSON: Scan Register Selection using Implicit State Enumeration\cite{KallaPartialScan}}
\label{alg:SIMPSON}
\end{algorithm}
\DecMargin{1em}

Note that the former part of this algorithm describes a breath-first-search (BFS) traversal in
state space. The traversal algorithm is a simple variation of BFS algorithm in graph theory where 
states are nodes and transitions are arcs. Each state is uniquely encoded by a combination
of a set of register data, which is usually represented by a Boolean vector.

Since a typical sequential circuit usually contains an functionally independent combinational logic
component, the traversal algorithm analyzes the combinational logic and derive the transition 
function for one-step reachability within current time-frame, and extend the result to complete 
execution paths through unrolling. If each state encoding (i.e. exact values in the selected registers) is
explicitly exposed and counted during the unrolling 
procedure, this unrolling is called {\bf explicit unrolling}. In the SIMPSON algorithm, the states cannot be 
directly read in the execution; instead, they are implicitly represented using a conjunction of several 
Boolean formulas. Such techniques differs from explicit unrolling are called {\bf implicit unrolling}.

However, the SIMPSON algorithm proposed by the author is usually not practical. The conjunctions of 
Boolean formulas are stored as BDDs, which is a canonical and convenient structure. Nevertheless, 
the size of BDD is easy to explode when the formulas become too long and too complicated. In the paper,
the author make a compromise between accuracy and cost, and turn to a partial scan technique rather than 
a full scan. In my research work, the aim is to explore a word-level technique which can make a full scan 
available.

\subsection{Word-level Data Flow on Modern Datapath Designs}
The integrity of modern digital circuit designs is very high. For example, processor A10 designed by Apple
integrates 3.3 billion transistors on a $125~mm^2$ chip \cite{AppleA10}. Such high integrity make the silicon 
implementation of large datapaths possible. In recent decades, leading by top-notch central processing units (CPUs)
and high bandwidth memory (HBM), 64-bit or even larger datapaths frequently appear in digital ICs.
Meanwhile, with the development of electrical design automation (EDA) tools, data flow on those datapaths
are directly described by the designer as word-level specifications. Therefore, it will be straightforward 
and beneficial for users if formal verification tools can work on word-level. Moreover, adopting word-level 
techniques will greatly reduce the state space and make verification executes faster by orders of magnitude.

@@

\begin{Example}[Verification of sequential multiplication hardware]

% fig:MIPS

Figure \ref{fig:MIPS} depicts a sequential multiplication hardware implementation within a 64-bit 
MIPS. Initially, one multiplicand is preloaded to the lower 64 bits of the product registers. 
Iteratively, the last significant bit (LSB) of current (temporary) product is used as flag to activate the ALU 
to add on the other multiplicand. For each iteration the data in product registers shift right by 1 bit.
Finally when the most significant bit (MSB) of preloaded multiplicand arrives the MSB of product registers,
the registers contains the result -- 128 bits product. The behavior can be concluded as following algorithm:

@@

We can see that when checking a single part of the whole structure, the number of clauses 
needed will increase to $k+1$ when using $k$-bit datapath. Considering the formula representing 
carry-in will become longer and longer, the final conjunction of all clauses will contain 
$O(2^k)$ Boolean operators. If by some means we can write the specification with only 3 variables
\begin{equation}
\label{eqn:wordspec}
S=P_{1/2}+B
\end{equation}
The state space will be greatly reduced.
\end{Example}

On the other hand, the user of conventional model checker will have to write all these clauses.
These clauses contain cross-literals, e.g. $s_2$ may associate with $a_1, p_0$, etc. If the user 
is not familiar with the implementation of this adder, those cross-literals will bring inconveniences
and confusions. However, if word-level techniques allow specification like equation \ref{eqn:wordspec},
the verification tool will be very user-friendly and straightforward.

\subsection{Prerequisites of Word-level Techniques}
When given a bit-level netlist, the prerequisites to use word-level techniques are to convert bit-level 
to word-level first. This conversion is usually completed by abstraction techniques.

An old but universally effective abstraction method is {\bf Lagrange's interpolation}. Instead of using 
it in real algebra, it can also be extended to Galois field. Here we use an example to illustrate the 
conversion in Galois field using Lagrange's interpolation.

@@

For each element in $\F_{2^3}$, we write down the truth table as follows:
\begin{table}[H]
\label{tab:truthtable}
\centering{\small
\begin{tabular}{|c|ccc|c|}
\hline
$\{a_2a_1a_0\}\in\mathbb{B}^3$   & $A\in \F_{2^3}$ &$\rightarrow$& $\{z_2z_1z_0\}\in\mathbb{B}^3$  & $Z\in \F_{2^3}$ \\
\hline
000  &0 &$\rightarrow$&000 & 0 \\
001  &1 &$\rightarrow$&001 & 1 \\
010  &$\alpha$ & $\rightarrow$ & 111& $\alpha^2 + \alpha + 1$ \\
011  &$\alpha + 1$ &$\rightarrow$& 111 & $\alpha^2 + \alpha + 1$ \\
100  &$\alpha^2$ &$\rightarrow$& 101 &  $\alpha^2+1$ \\
101  &$\alpha^2 + 1$ &$\rightarrow$&011 & $\alpha+1$ \\
110  &$\alpha^2 + \alpha$&$\rightarrow$& 101 &$\alpha^2 + 1$ \\
111  &$\alpha^2 + \alpha + 1$ &$\rightarrow$& 101 &$\alpha^2 + 1$\\
\hline
\end {tabular}
}
\caption{Truth table for mappings in $\mathbb{B}^3$ and $\F_{2^3}$}
\end{table}
Now our objective is to abstract a function over GF $\F_{2^3}$ about word-level variables, i.e. $Z = \Func(A)$.
Recall the definition of Lagrange's interpolation:
\begin{equation}
\label{eqn:Lagrange}
\Func(x) =  \sum_{k=1}^N \left[ \prod_{(0\leq j \leq k-1),(j\neq i)}\frac{x-x_j}{x_i-x_j} \cdot y_k \right]
\end{equation}
The geometry meaning of Lagrange's interpolation in real algebra is: given $N$ dots with coordinates $(x_i,y_i)$,
they can always be fitted into a polynomial function with at most $N-1$ degree, and that function can be 
written in the form of equation \ref{eqn:Lagrange}. In this example, although defined in Galois field instead of 
real number field, the essential concept of Lagrange's interpolation keeps the same. We can get 8 ``dots":
\begin{align*}
Uni-form&: (a_2\alpha^2+a_1\alpha+a_0,z_2\alpha^2+z_1\alpha+z_0) \gets (A,Z) \\
Dot~1&: (0, 0) \gets (000,000) \\
Dot~2&: (1,1) \gets (001,001) \\
Dot~3&:  (\alpha,\alpha^2 + \alpha + 1) \gets (010,111)\\
Dot~4&:  (\alpha + 1,\alpha^2 + \alpha + 1)\gets (011,111) \\
Dot~5&:  (\alpha^2,\alpha^2+1) \gets (100,101)\\
Dot~6&:  (\alpha^2 + 1,\alpha+1)\gets (101,011)\\
Dot~7&:  (\alpha^2 + \alpha,\alpha^2 + 1)\gets (110,101)\\
Dot~8&:  (\alpha^2 + \alpha + 1,\alpha^2 + 1)\gets (111,101)
\end{align*}
Substitute 8 $(x_i,y_i)$ pairs in equation \ref{eqn:Lagrange} with these 8 ``dots" in $\F_{2^3}$.
The result is a polynomial function with degree no greater than 7:
\begin{align*}
Z =& \Func(A) \\
=& (\alpha^2+\alpha+1) A^7 + (\alpha^2+1) A^6 + \alpha A^5 + (\alpha+1) A^4 \\
&+ (\alpha^2+\alpha+1)A^3+ (\alpha^2+1)A
\end{align*}

@@

The Lagrange's interpolation theorem also proves the existence of a word-level abstraction for a bit-level netlist.
In practice, Lagrange's interpolation is too weak on scalability. Our approach uses abstraction based on 
Gr\"obner basis with abstraction term order (ATO), which is briefly introduced in preliminaries.

\section{FSM Reachability using Algebraic Geometry}
\label{sec:reach}
% \subsection{An example illustrating our proposed approach}
We use symbolic state reachability with algebraic
geometry concepts. It is an abstraction based on word operand
definition of datapaths in circuits, and it can be applied
to arbitrary FSMs by bundling a set of bit-level variables together as
one or several word-level variables.  The abstraction polynomial,
encoding the reachable state space of the FSM, is obtained through
computing a GB over $\Fkk$ of the polynomials of the circuit using an
elimination term order based on Theorem \ref{thm:elim}.

@@

The {\it transition function} $\Delta$ is given by Boolean equations
  of the flip-flops of the circuit: $t_i = \Delta_i(s, x)$, where
  $t_i$ is a next state variable, $s$ represents the present   state
  variables and $x$ represents the input variables. The {\it
    transition relation of the FSM} is then represented as:  
\begin{equation} 
T(s, x, t) =   \prod_{i=1}^{n} (t_i \overline{\oplus } \Delta_i)
\end{equation}
where $n$ is the number of flip flops, and $\overline{\oplus}$ is XNOR
operation. Let $from$ denote the set of initial states, then the
image of the initial states, under the transition function $\Delta$ is
finally computed as:
\begin{align}
\label{img}
to = \text{Img}(\Delta, from) = \exists _s ~\exists _x ~[ T(s, x, t)
  \cdot from ] 
%= \exists _s ~\exists _x ~\prod_{i=1}^{n} (t_i \overline{\oplus } \Delta_i)\cdot from
\end{align}

Here, $\exists x (f)$ represents the {\it existential quantification
  of $f$ w.r.t. variable $x$}.

@@

Let us describe the application of the algorithm on the  FSM circuit
of figure \ref{fig:fsm}. {\it We will first describe its operation at the
Boolean level, and then describe how this algorithm can be implemented
using algebraic geometry at word level.}

@@

In Line 1 of the BFS algorithm, assume that the initial state
is $S_3$ in figure \ref{fig:fsm}(b), which is encoded as 
$S_3 = \{11\}$. Using Boolean variables $s_0, s_1$ for the present
states, $from^0 = s_0\cdot s_1$ is represented as a Boolean formula.

\begin{Example}
For the circuit in figure \ref{fig:fsm} (b), we have the transition
functions of the machine as:

@@

When the formula of Eqn. (\ref{img}) is applied to compute 1-step
reachability, $to = \exists _{s_0, s_1, x} (\Delta_1 \cdot \Delta_2
\cdot from^0)$, we obtain $to = \overline{t_0}\cdot t_1$, which denotes
the state $S_1 = \{01\}$ reached in 1-step from $S_3$.
In the next iteration, the algorithm uses state $S_1 = \{01\}$ as the
current (initial) state, and computes $S_2 = \{10\} = t_0\cdot
\overline{t_1}$ as the next reachable state, and so on. 
\end{Example}

@@

\subsection{FSM Traversal at word-level over $\Fkk$} 
The state transition graph (STG) shown in figure \ref{fig:fsm}(a) uses a
2-bit Boolean vector to represent 4 states $\{S_0, S_1, S_2,
S_3\}$. We map these states to elements in $\mathbb{F}_{2^2}$, where
$S_0 = 0, S_1 = 1, S_2 = \alpha, S_3 = \alpha+1$. Here, we take 
$P(X) = X^2+X+1$ as the irreducible polynomial to construct
$\mathbb{F}_4$, and $P(\alpha) = 0$ so that $\alpha^2 + \alpha + 1 =
0$.

{\it Initial state:} In Line 1 of algorithm \ref{alg:BFS}, the initial state is
specified by means of a corresponding polynomial $f = \mathcal{F}(S) =
S - 1 - \alpha$. Notice that if we consider the ideal generated by the
initial state polynomial, $I = \langle f\rangle$, then its variety
$V(I) = 1+\alpha$ corresponds to the state encoding $S_3 = \{11\} =
1+\alpha$, where a polynomial in word-level variable $S$ encodes the
initial state. 
% with only one generator $f$, its variety
% $V(I) = \{\gamma\ |\ \gamma \in \mathbb{F}_{2^2}, \gamma = 1+\alpha\}$, which equals to $\{1+\alpha\}$, the only
% valid value $S_3$ can take.

@@

{\bf Set operations:} In Lines 5 and 6 of
algorithm  \ref{alg:BFS}, we need  \textbf{union}, 
\textbf{intersection} and \textbf{complement} of varieties over
$\mathbb{F}_{2^k}$, for which we again use algebraic geometry concepts.

@@

Assume we have $k$-bit datapath and $n$-bit PIs. In Galois field, we need to carefully partition $n$ PIs
such that states of each partition can be covered by a univariate polynomial respectively.

>>>>

\chapter{Word-level Traversal of Finite State Machines using Algebraic Geometry}
\label{ch:reacha}
Reachability analysis is a basic component of sequential circuit verification, especially for 
formal equivalence checking and model checking techniques. Concretely, in modern synthesis 
tools, in order to improve various performance indicators such as latency, clock skew or power
density, sequential optimization techniques such as retiming \cite{retiming}, scan logic \cite{scan},
sequential power optimization \cite{mathur2009power} and clock-gating techniques \cite{clockgating}.
% some major refactoring and attachments are made on original designs. 
These modifications may introduce malfunctions to the original logic and cause problems. 
Based on traditional localized simulation or formal verification method (e.g. equivalence checking), 
designers are reluctant to make aggressive optimization
since the malfunctions are considered as ``faults" in this circuit.
However, if the circuit behavior is carefully investigated, it may become evident that 
those ``faults" will never be activated/excitated during a restricted execution starting 
from legal initial states and with legal inputs. Thus we will call those ``faults" as 
{\bf spurious faults} (false negatives), since they will not affect the circuit's normal behavior.

Almost all practical sequential logic components can be modeled as finite state machines (FSMs). 
If we apply constraints upon the machine to make it start from designated initial states, and 
take specific legal inputs, a set of reachable states can be derived. 
As long as the ``faults" can be modeled as ``bad states", we can judge whether they are 
spurious faults by checking if they belong to the reachable states. From the spurious fault validation 
perspective, reachability analysis is a must when developing full set of sequential circuit verification
techniques.

There are quite a few methods to perform reachability checking on FSMs. One among them is 
state space traversal. Conventionally the algorithm is based on bit-level techniques such as
binary decision diagrams (BDDs) and Boolean logic. We propose a new traversal algorithm on word-level,
which brings critical advantages. In this chapter the approach will be described and discussed in depth, 
with examples and experiments showing its feasibility when applied on general circuit benchmarks.

\section{Motivation}
Before introducing the details of our approach, there are a couple of questions to ask: why is it
necessary to execute word-level FSM traversal? Why does algebraic geometry make this happen?
The answers can be found in this section, as statement of the motivation of our research.

\subsection{FSM Traversal Algorithms}
Sequential circuits are modeled as FSMs, which can be implemented as graph. Thus a 
graph-traversal based algorithm is created to analysis the reachable states \cite{coudert2003unified}.
A traversal algorithm using concept ``implicit state enumeration" is proposed \cite{cho1993redundancy}.
Concretely, the algorithm is written as follows:

\begin{algorithm}[H]
\SetAlgoNoLine
\LinesNumbered
 \KwIn{Sequential circuit with given number of registers}
 \KwOut{Scan registers listed in decreasing order of their non-controllability}

  $from^0 = reached = S^0$\;
  $i = 0$\;
  \While{TRUE}
  {
  	$i++$\;
	$to^i = $IMAGE$(\Delta,from^{i-1})$\;
	$new^i = to^i \cdot \overline{reached}$\;
	\For{each state variable $r_j$}
	{
		record\_if\_transitions\_present\_or\_missing$(r_j,new^i)$\;
		compute\_degree\_of\_unsettability$(r_j,new^i)$\;
	}
	\If{$new^i == 0$}
	{
		break\;
	}
	$reached = reached + new^i$\;
	$from^i = new^i$\;
  }\
\Return{$reached$}
\caption {FSM Traversal using Implicit State Enumeration\cite{KallaPartialScan}}
\label{alg:SIMPSON}
\end{algorithm}
\DecMargin{1em}

Above algorithm describes a breath-first-search (BFS) traversal in
state space. The traversal algorithm is a simple variation of BFS algorithm where 
states are nodes and transitions are arcs. Each state is uniquely encoded by a combination
of a set of register data, which is usually represented by a Boolean vector.

Since a typical sequential circuit usually contains a combinational logic
component, the traversal algorithm analyzes the combinational logic and derives the transition 
function for one-step reachability within current time-frame, and extends the result to complete 
execution paths through unrolling. If each state encoding (i.e. exact values in the selected registers) is
explicitly analyzed and counted during the unrolling 
procedure, this unrolling is called {\bf explicit unrolling}. In the BFS traversal algorithm, the states cannot be 
directly read in the execution; instead, they are implicitly represented using a conjunction of several 
Boolean formulas. Such techniques differs from explicit unrolling are called {\bf implicit unrolling}.

However, the BFS traversal algorithm proposed by the author is usually not practical. The conjunctions of 
Boolean formulas are stored as BDDs, which is a canonical and convenient structure. Nevertheless, 
the size of BDD explodes when the formulas become too long and too complicated. In \cite{cho1993redundancy},
the authors make a compromise between accuracy and cost, and turn to approximate reachability. 
In this research work, the aim is to explore a word-level technique which can make a accurate reachability
analysis available.

\subsection{Word-level Data Flow on Modern Datapath Designs}
The level of integration of modern digital circuit designs is very high. For example, processor A10 designed by Apple
integrates 3.3 billion transistors on a $125~mm^2$ chip \cite{AppleA10}. Such a high density makes the silicon 
implementation of large datapaths possible. In recent decades, 64-bit or even larger datapaths frequently appear in 
modern digital IC designs such as powerful central processing units (CPUs) and high bandwidth memory (HBM).
Meanwhile, with the development of electronic design automation (EDA) tools, data flow 
is described by the designer as word-level specifications. Therefore, it will be straightforward 
and beneficial for users if formal verification tools can work on word-level. Moreover, adopting word-level 
techniques will greatly reduce the state space and make verification more efficient.

@@

\begin{Example}

\begin{figure}[H]
\centering{
\includegraphics[width=\textwidth]{newfig/MIPS.pdf}
\caption{A 64-bit sequential multiplication hardware design}
\label{fig:MIPS}}
\end{figure}

Figure \ref{fig:MIPS} depicts a sequential multiplication hardware implementation within a 64-bit 
MIPS. Initially, one multiplicand is preloaded to the lower 64 bits of the product registers. 
Iteratively, the last significant bit (LSB) of current (temporary) product is used as flag to activate the ALU 
to add on the other multiplicand. For each iteration the data in product registers shifts right by 1 bit.
Finally when the most significant bit (MSB) of preloaded multiplicand arrives at the MSB of product registers,
the registers contains the result -- 128 bits product. The behavior can be described by the following algorithm:

% \vspace{0.2in}

@@

We can see that when checking a single part of the whole structure, the number of clauses 
needed will increase to $k+1$ when using $k$-bit datapath. Considering the formula representing 
carry-in will become longer and longer, the final conjunction of all clauses will contain 
$O(2^k)$ Boolean operators. If by some means we can write the specification with only 3 variables
\begin{equation}
\label{eqn:wordspec}
S=P_{1/2}+B
\end{equation}

Since the abstraction to word-level reduces symbolic storage and execution cost, the complexity 
to traverse the state space is greatly reduced.
\end{Example}

On the other hand, when implementation details of the datapath are not available, it is not convenient 
any more for users of conventional model checker. The reason is that the user has to write all clauses 
for the implementation, which contains cross-literals, e.g. $s_2$ may associate with $a_1, p_0$, etc. If the user 
is not familiar with the implementation of this adder, those cross-literals will bring
confusions. However, if word-level techniques allow specification like Equation \ref{eqn:wordspec},
the verification tool will be very user-friendly and straightforward even if the implementation details
are in a black box.

\subsection{On the Existence of Word-level Abstraction}
When given a bit-level netlist, the prerequisites to use word-level techniques are to convert bit-level 
to word-level first. This conversion is usually completed by abstraction techniques.

An old but universally effective abstraction method is {\it Lagrange's interpolation}. Instead of using 
it in real algebra, it can also be extended to Galois field. Here we use an example to illustrate the 
conversion in Galois field using Lagrange's interpolation.

@@

For each element in $\F_{2^3}$, we write down the truth table as follows:
\begin{table}[H]
\label{tab:truthtable}
\centering{\small
\begin{tabular}{|c|ccc|c|}
\hline
$\{a_2a_1a_0\}\in\mathbb{B}^3$   & $A\in \F_{2^3}$ &$\rightarrow$& $\{z_2z_1z_0\}\in\mathbb{B}^3$  & $Z\in \F_{2^3}$ \\
\hline
000  &0 &$\rightarrow$&000 & 0 \\
001  &1 &$\rightarrow$&001 & 1 \\
010  &$\alpha$ & $\rightarrow$ & 111& $\alpha^2 + \alpha + 1$ \\
011  &$\alpha + 1$ &$\rightarrow$& 111 & $\alpha^2 + \alpha + 1$ \\
100  &$\alpha^2$ &$\rightarrow$& 101 &  $\alpha^2+1$ \\
101  &$\alpha^2 + 1$ &$\rightarrow$&011 & $\alpha+1$ \\
110  &$\alpha^2 + \alpha$&$\rightarrow$& 101 &$\alpha^2 + 1$ \\
111  &$\alpha^2 + \alpha + 1$ &$\rightarrow$& 101 &$\alpha^2 + 1$\\
\hline
\end {tabular}
}
\caption{Truth table for mappings in $\mathbb{B}^3$ and $\F_{2^3}$}
\end{table}
Now our objective is to abstract a function over GF $\F_{2^3}$ about word-level variables, i.e. $Z = \Func(A)$.
Recall the definition of Lagrange's interpolation:
\begin{equation}
\label{eqn:Lagrange}
\Func(x) =  \sum_{k=1}^N \left[ \prod_{(0\leq j \leq k-1),(j\neq i)}\frac{x-x_j}{x_i-x_j} \cdot y_k \right]
\end{equation}
The geometric meaning of Lagrange's interpolation in real algebra is: given $N$ dots with coordinates $(x_i,y_i)$,
they can always be fitted into a polynomial function with at most $N-1$ degree, and that function can be 
written in the form of equation \ref{eqn:Lagrange}. In this example, although defined in a Galois field instead of 
the real number field, the essential concept of Lagrange's interpolation remains the same. 
We can get 8 points in the affine space:
\begin{align*}
\text{Generic form}&: (a_2\alpha^2+a_1\alpha+a_0,z_2\alpha^2+z_1\alpha+z_0) \gets (A,Z) \\
\text{Point }1&: (0, 0) \gets (000,000) \\
\text{Point }2&: (1,1) \gets (001,001) \\
\text{Point }3&:  (\alpha,\alpha^2 + \alpha + 1) \gets (010,111)\\
\text{Point }4&:  (\alpha + 1,\alpha^2 + \alpha + 1)\gets (011,111) \\
\text{Point }5&:  (\alpha^2,\alpha^2+1) \gets (100,101)\\
\text{Point }6&:  (\alpha^2 + 1,\alpha+1)\gets (101,011)\\
\text{Point }7&:  (\alpha^2 + \alpha,\alpha^2 + 1)\gets (110,101)\\
\text{Point }8&:  (\alpha^2 + \alpha + 1,\alpha^2 + 1)\gets (111,101)
\end{align*}
Substitute 8 $(x_i,y_i)$ pairs in Equation \ref{eqn:Lagrange} with these 8 points in $\F_{2^3}$.
The result is a polynomial function with degree no greater than 7:
\begin{align*}
Z =& \Func(A) \\
=& (\alpha^2+\alpha+1) A^7 + (\alpha^2+1) A^6 + \alpha A^5 + (\alpha+1) A^4 \\
&+ (\alpha^2+\alpha+1)A^3+ (\alpha^2+1)A
\end{align*}

@@

The Lagrange's interpolation theorem also proves the existence of a word-level abstraction for a bit-level netlist.
In practice, Lagrange's interpolation is not scalable. 
The reason is that it needs entire function (state space), but usually we only have the circuit representation
of the FSM. Considering this fact, a symbolic method is needed.
Our approach in this section uses abstraction based on 
Gr\"obner basis with abstraction term order (ATO), which is briefly introduced in Section \ref{sec:abstraction}.

\subsection{Significance of Developing Word-level FSM Traversal}
At the end of this section, we summarize each subsection and conclude the motivation of our research.
First, we start from the basic FSM traversal algorithms and exhibit their
difficulties to overcome. Secondly, we state the observation that word-level description is 
increasingly important and common in characterizing the data flow on modern large-size datapaths. 
Last but not least, we give instances that some prerequisites for supporting word-level techniques 
are already available.

To conclude, first we state the importance of performing FSM traversal in sequential circuit reachability analysis.
To overcome the vast cost brought by searching in a large space, we propose to use word-level 
polynomials to represent the states and transition relations. As a result, states are categorized into 
sets represented by a small number of word-level polynomials (more specifically, the varieties to the polynomial ideals),
and multiple transition relations are therefore merged together. All of these efforts reduces the space cost of 
state space, meanwhile lower the time complexity to traverse such a state space. By using Lagrange's interpolation,
we prove the feasibility of word-level techniques on general circuits' verification.
Thus we cover both the necessity and sufficiency of developing such 
a word-level traversal technique in our work.
They are also answers to the questions at the beginning of this section.

\section{FSM Reachability using Algebraic Geometry}
\label{sec:reach}
% \subsection{An example illustrating our proposed approach}
We use symbolic state reachability with algebraic
geometry concepts. It is an abstraction based on word operand
definition of datapaths in circuits, and it can be applied
to arbitrary FSMs by bundling a set of bit-level variables together as
one or several word-level variables.  The abstraction polynomial,
encoding the reachable state space of the FSM, is obtained through
computing a GB over $\Fkk$ of the polynomials of the circuit using an
elimination term order based on Theorem \ref{thm:elimth}.

\subsection{FSM model for sequential circuits}
A finite state machine (FSM) is a mathematical model of computation for designing and analyzing sequential logic 
circuits. If a FSM's primary outputs depend on primary inputs and present state inputs, it is named as a \textit{Mealy machine};
the formal definition is as follows:
\begin{Definition}
A Mealy machine is an $n$-tuple $\mathcal M = (\Sigma,O,S,S^0,\Delta,\Lambda)$ where
\begin{itemize}
\item $\Sigma$ is the input label, $O$ is the output label;
\item $S$ is the set of states, $S^0\subseteq S$ is the set of initial states;
\item $\Delta:\ S\times\Sigma\to S$ is the next state transition function;
\item $\Lambda:\ S\times\Sigma\to O$ is the output function.
\end{itemize}
\end{Definition}
The other kind of FSM is \textit{Moore machine}, its difference from Mealy machine is that
its primary outputs only depend on the present states, i.e. the output function is defined as
$$\Lambda:\ S \to O$$
Typical sequential circuits can be depicted as Figure \ref{fig:seqmodel}(a). Primary inputs
$x_1,\dots,x_m \in \Sigma$, and primary outputs $z_1,\dots,z_n\in O$. Signals $s_1,\dots,s_k$ 
are present state (PS) variables, $t_1,\dots,t_k$ are next state (NS) variables.
We can define 2 $k$-bit words denoting the PS/NS variables as there are $k$ flip-flops
in the datapath: $S = (s_1,\dots,s_k), ~T=(t_1,\dots,t_k)$. Transition function
at bit level are defined as $\Delta_i$: 
$$t_i = \Delta_i(s_1,\dots,s_k,x_1,\dots,x_m)$$
\begin{figure}[H]
\centering{
%\begin{minipage}{12cm}
\includegraphics[width=5.5in]{newfig/seqmodel.eps}
% \vspace{-0.2in}
\caption{FSM models of sequential circuits}
%\end{minipage}
\label{fig:seqmodel}}
\end{figure}
In some cases, arithmetic computations are implemented as Moore machines where input operands
are loaded into register files $R$ and the FSM is executed for $k$ clock cycles.
We can simplify them to the model in Figure \ref{fig:seqmodel}(b).

@@

The {\it transition function} $\Delta$ is given by Boolean equations
  of the flip-flops of the circuit: $t_i = \Delta_i(s, x)$, where
  $t_i$ is a next state variable, $s$ represents the present   state
  variables and $x$ represents the input variables. The {\it
    transition relation of the FSM} is then represented as:  
\begin{equation} 
T(s, x, t) =   \prod_{i=1}^{n} (t_i \overline{\oplus } \Delta_i)
\end{equation}
where $n$ is the number of flip flops, and $\overline{\oplus}$ is XNOR
operation. Let $from$ denote the set of initial states, then the
image of the initial states, under the transition function $\Delta$ is
finally computed as:
\begin{align}
\label{eqn:img}
to = \text{Img}(\Delta, from) = \exists _s ~\exists _x ~[ T(s, x, t)
  \cdot from ] 
%= \exists _s ~\exists _x ~\prod_{i=1}^{n} (t_i \overline{\oplus } \Delta_i)\cdot from
\end{align}

Here, $\exists x (f)$ represents the {\it existential quantification
  of $f$ w.r.t. variable $x$}. In Boolean logic, this operator is implemented as
  $$\exists x (f) = f_x\lor f_{\overline{x}}$$

@@

Let us describe the application of the algorithm on the  FSM circuit
of Figure \ref{fig:fsm}. {\it We will first describe its operation at the
Boolean level, and then describe how this algorithm can be implemented
using algebraic geometry at word level.}

@@

In Line 1 of the BFS algorithm, assume that the initial state
is $S_3$ in Figure \ref{fig:fsm}(b), which is encoded as 
$S_3 = \{11\}$. Using Boolean variables $s_0, s_1$ for the present
states, $from^0 = s_0\cdot s_1$ is represented as a Boolean formula.

\begin{Example}
For the circuit in Figure \ref{fig:fsm} (b), we have the transition
functions of the machine as:

@@

When the formula of Equation \ref{eqn:img} is applied to compute 1-step
reachability, $to = \exists _{s_0, s_1, x} (\Delta_1 \cdot \Delta_2
\cdot from^0)$, we obtain $to = \overline{t_0}\cdot t_1$, which denotes
the state $S_1 = \{01\}$ reached in 1-step from $S_3$.
In the next iteration, the algorithm uses state $S_1 = \{01\}$ as the
current (initial) state, and computes $S_2 = \{10\} = t_0\cdot
\overline{t_1}$ as the next reachable state, and so on. 
\end{Example}

@@

\subsection{FSM Traversal at word-level over $\Fkk$} 
The state transition graph (STG) shown in Figure \ref{fig:fsm}(a) uses a
2-bit Boolean vector to represent 4 states $\{S_0, S_1, S_2,
S_3\}$. We map these states to elements in $\mathbb{F}_{2^2}$, where
$S_0 = 0, S_1 = 1, S_2 = \alpha, S_3 = \alpha+1$. Here, we take 
$P(X) = X^2+X+1$ as the irreducible polynomial to construct
$\mathbb{F}_4$, and $P(\alpha) = 0$ so that $\alpha^2 + \alpha + 1 =
0$.

{\it Initial state:} In Line 1 of Algorithm \ref{alg:BFS}, the initial state is
specified by means of a corresponding polynomial $f = \mathcal{F}(S) =
S - 1 - \alpha$. Notice that if we consider the ideal generated by the
initial state polynomial, $I = \langle f\rangle$, then its variety
$V(I) = 1+\alpha$ corresponds to the state encoding $S_3 = \{11\} =
1+\alpha$, where a polynomial in word-level variable $S$ encodes the
initial state. 
% with only one generator $f$, its variety
% $V(I) = \{\gamma\ |\ \gamma \in \mathbb{F}_{2^2}, \gamma = 1+\alpha\}$, which equals to $\{1+\alpha\}$, the only
% valid value $S_3$ can take.

@@

{\bf Set operations:} In Lines 5 and 6 of
Algorithm  \ref{alg:BFS}, we need  \textbf{union}, 
\textbf{intersection} and \textbf{complement} of varieties over
$\mathbb{F}_{2^k}$, for which we again use algebraic geometry concepts.

@@

% fig:PI

Assume we have $k$-bit datapath and $n$-bit PIs. In finite field, we need to carefully partition $n$ PIs
such that states of each partition can be covered by a univariate polynomial respectively.
